Spatial variation of the native colon microbiota in healthy adults

### 2017-10-31 submission/manuscript.Rmd
almost all of pat edits made

#### Random Forest classification models identify important Operational Taxonomic Units (OTUs) on each side
To identify OTUs that were distinct at each site, we constructed several Random Forest models trained using OTU relative abundances. We used 10-fold cross validation to build the first model to classify the luminal versus mucosal samples for the proximal and distal sides, independently (Figure S1A). The models performed well when classifying these samples (proximal AUC = 0.764, distal AUC = 0.908). The OTUs that were most predictive of each site were identified by their greatest mean decrease in accuracy when removed from the model. For distinguishing the proximal lumen and mucosa, OTUs affiliated with the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and _Enterobacteraceae_ families were included in the best model (Figure 4A). The model to differentiate between the distal lumen and mucosa included OTUs affiliated with the _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and _Anaerococcus_ families (Figure 4B). These results indicated that there were fine differences between the different sites of the colon, and that these could be traced to specific OTUs on each side.

Next, we built a Random Forest model to differentiate the proximal and distal lumenal samples using 10-fold cross validation. The model performed best when distinguishing the proximal versus distal mucosa (Figure S1B, AUC = 0.850) compared to the proximal versus distal lumen (AUC = 0.580). OTUs that were differentially abundant between the distal and proximal mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ genera (Figure 5A). Differentially abundant OTUs of the proximal and distal lumen included three OTUs of the _Bacteroides_ genus, a _Clostridium IV_ OTU and an _Oscillibacter_ OTU (Figure 5B). This analysis found that some of the same OTUs that are distinct between the mucosa and lumen also helped to differentiate between the two sides- such as _Anaerococcus_ and _Finegoldia_.

#### Bacterial OTUs associated with cancer are found in healthy individuals 
Given that specific bacterial species have been associated with colorectal cancer and IBD, we probed our sample set for these OTUs. Among our 100 samples, the most frequent sequence associated with the _Fusobacterium_ genus was OTU179, which aligns via BLASTn to _Fusobacterium nucleatum subsp animalis_ (100% over full length). This is the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors. @Castellarin2011. There were 14 patient samples with the _F. nucleatum subsp. animalis_ sequences. Of the samples with the highest abundance of _F. nucleatum subsp. animalis_, four of the samples were from the proximal mucosa and three from the distal mucosa (Figure S2A). The second most frequent _Fusobacterium_ sequence was OTU472, which aligned with 99% identity to _F. varium_. In addition to _F. nucleatum_, _F. varium_ has been associated with IBD @Lee2016. Four study participants harbored _F. varium_ and the samples were split evenly between the proximal and distal mucosa (Figure S2B). OTU152 was similar to the members of the _Porphyromonas_ genus and the most frequent sequence in that OTU aligned to _Porphyromonas asacharolytica_ (99% over full length), another bacterium commonly detected and isolated from colorectal tumors. OTU152 was only detected on the distal mucosa, and in fact was one of the OTUs the classification model identified as separating distal and proximal sides (Figure S2C). Among the 11 distal mucosa samples that were positive for _P. asacharolytica_, the relative abundances for this OTU ranged from 0.01% - 16%. Thus, disease-associated OTUs could be found in our sample set of 20 healthy individuals.

### Discussion
Here we identified bacterial taxa that were specific to the lumen and mucosa of the proximal and distal sides of the human colon from samples collected during unprepared colonoscopy. We found that all locations contained a range of phyla abundances and a range of diversity, but that there was a wide variability between subjects. Pairwise comparisons of each of the sites revealed that the proximal mucosa and lumen were most similar to each other. Further, comparison of colonoscopy-collected samples with samples collected from stool at home showed that the distal lumen was most similar to feces. Random Forest models built on OTU relative abundances from each sample identified microbes that were particular to each location of the colon. Finally, we were able to detect some bacterial OTUs associated with colonic disease in our healthy patient cohort. Using unprepped colonoscopy and machine learning, we have identified bacterial taxa specific to the healthy proximal and distal human colon.

When examining the relative abundance of the dominant phyla at each site (i.e. _Bacteriodes_ and _Firmicutes_), there was a wide amount of variation. This likely reflects not only the variability between human subjects, casued by differences in age, sex, diet, but also biogeogrphical "patchiness" in the gut microbiome. Several studies have noted that the bacteria recoverable from the same mucosal sample location can be vastly different when the samples are taken just 1 cm away from each other @Hong2011. Similar patchiness is also observed in lumenal contents and fecal samples themselves; there is observed separation of different interacting microbes along the length of a stool sample, for instance @Stearns2011. That said, across our samples the mucosal samples harbor more _Proteobacteria_, consistent with previous studies comparing mucosal swabs to lumenal content in humans @Albenberg2014. Hence, the conclusions we can draw from phyla analysis may be impacted by patchiness between subjects.

To get around the noisiness from a diverse set of samples, we built a Random Forest model to identify microbes specific to each side. For each comparison we identified the top five OTUs that were strongly predictive of one site or another. Generally, OTUs identified in each location were consistent with known physiological gradients along the gut axis @Donaldson2015. For instance, the proximal mucosa contains the highest oxygen concentratons of the colon and  harbored mucosa-associated facultative anaerobes such as _Actinomyces_ and _Enterobacteraceae_ and aerobic _Psuedomonas_. The distal mucosa was far more likely to host strictly anaerobic species such as _Porphyromonas_, _Anaerococcus_, _Finegoldia_ and _Peptoniphilus_. Thus the gut microenvironment of each location likely enriches for these specific microbes.

In addition to identifying features that are specific to each side of the gut, the ability of the Random Forest to classify samples can serve as a proxy for similarity. That is, a higher AUC value indicates the samples are more efficiently classfied (and thus more different) than a model with a lower AUC value. For instance, the model separating the proximal and distal mucosa has an AUC of 0.850 whereas the model for classifying the proximal and distal lumen has a much lower AUC of 0.580. Further, the latter model required 44 OTUs to best separate the samples whereas the models separating the mucosa only needed 10 OTUs. The much lower AUC and need for a high number of features compared to other models suggest these locations are the most similar of the comparisons tested. We speculate that the model was least effective at classifying the proximal and distal lumenal contents because the samples are arguably composed of the same bacteria but differ in water content.

We detected _F. nucleatum_ and _P. asacharolytica_ in 8 and 5 of our subjects, respectively. These bacteria have been shown to be predictive of colorectal cancer in humans @Baxter2016 and have oncogenic properties in cell culture and in mice @Sears2014. Interestingly, while _F. nucleatum_ was found on both sides of the colon, _P. asacharolytica_ was only detected in the distal mucosa. Not much is known about the distribution of _P. asacharolytica_ but given its documented anaerobic characteristics and asacharolytic metabolism, it might not be surprising that it resides in the less-oygen-rich and proteinaceous distal mucosa @Albenberg2014. In studies examining bacteria on colorectal cancer tumors, _F. nucleatum_ is more commonly detected on proximal-sided tumors, and distribution of _F. nucleatum_ decreases along the colon to rectum @Mima2016.  Of the 8 (40%) individuals positive for _F. nucleatum_ in this study, the bacterium was spread across the proximal mucosa, distal lumen and distal mucosa. Data examining bacterial biofilms on the mucosa of CRC tumors suggests that _Fusobacteria_ species are more commonly found on proximal tumors and in biofilms, indicating that it is not only the presence of the bacteria but the organization of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis @Dejea2014. Finally, _Fusobacterium_ and _Porphyromonas_ species have been known to not only co-occur on CRC tumors but also to synergistically promote tumorigenesis in an oral cancer model @Whitmore2014 @Flynn2016. Thus, further analysis of the distribution and activities of these pathogens may elucidate a mechanism for development of IBD or CRC subtypes in the proximal or distal colon.

The _Fusobacterium_ species _nucleatum_ and _varium_ have been commonly isolated from mucosal biopsies of patients with IBD @Lee2016. Laboratory experiments with these isolates have shown that disease-isolated _F. nucleatum_ are more invasive and stimulate more TNF-$\alpha$ production than strains from healthy individuals @Strauss2011, suggesting the bacteria may increase inflammation in the gut as well @Dharmani2011. _F. varium_ isolated from UC patients caused colonic ulcers in an experimental mouse model @Ohkusa2003. _F. varium_ was only detected in three of our study participants and two of those samples were isolated form the proximal mucosa (Figure S2B).  _F. varium_ is most commonly isolated from UC patient biopsies from the ileum or cecum @Ohkusa2002, suggesting this species may exhibit preference for the different environmental conditions of these gastrointestinal sites. Further work will assess how gut environment may select for species which may then cause localized disease.

Specific comparisons of our findings to previously published gut biogeography studies are confounded by the use of bowel preparation methods in most other studies. A rare report of a matched-colonoscopy study sampled 18 patient's colonic mucosa and lumenal contents prior to and after bowel cleansing @Shobar2016. This study found that mucosal and lumenal samples were distinguishable prior to bowel cleansing, but that bowel preparation resulted in an increase in shared OTUs between each site @Shobar2016. After seven days, bowel cleansing not only made the samples harder to distinguish, but it also decreased in diversity across sites. Bowel preparation clearly biases into the microbes recovered from sampling the lumen or mucosa of a prepared bowel.

By revealing specific differences in microbial populations at each location in the gut via sampling an unprepared bowel, we can begin to form hypothesies about how specific host-microbe interactions can affect disease progression of proximal and distal CRC and IBD subtypes. To this point, 16S rRNA gene sequecing community profiling studies do not provide enough information to fully probe these questions. In particular, 16S sequencing cannot not profile the host characteristics at each site. Combining the unprepared colonoscopy approach with analysis of multi-omic sequencing data may be useful in further characterizing host-microbiome interactions along the gut axis for both health and disease.

@@

#### Sample collection
At a baseline visit, subjects were consented and given a home collection stool kit. At least one week prior to the scheduled colonoscopy, subjects were to collect whole stool at home and ship the samples to a research coordinator on ice. Notably, subjects did not undergo any bowel preparation method prior to sampling. On the procedure day, subjects reported to the Michigan Clinical Research Unit at the University of Michigan Health System. Patients were consciously sedated using Fentanyl, Versed and/or Benadryl as appropriate. A flexible sigmoidoscope was first inserted about 25cm into the colon and endoscopy brush used to collect lumenal/stool contents. Two lumenal samples were collected and the contents immediately deposited into RNAlater (Fischer) and flash-frozen in liquid nitrogen. The brushes were withdrawn and biopsy forceps were used to collect mucosal biopsies on sections of the colon that were pink and free of stool matter. Three mucosal biopsies were collected and flash-frozen in RNAlater. These samples comprised the distal or distal colon samples. The sigmoidoscope was then withdrawn and a pediatric colonoscope was inserted to reach the ascending colon. Samples were then collected as in the distal colon and the colonoscope withdrawn. All samples were stored at -80$^\circ$C.

#### Sample processing, sequencing and analysis
DNA extraction was performed using the PowerMicrobiome DNA/RNA Isolation Kit (MO BIO Laboratories). For tissue biopsies, Bond-Breaker TCEP solution (Fisher) and 2.8mm ceramic beads (MO BIO Laboratories) were added to the bead beating step to enhance DNA recovery from mucosal samples. The resulting DNA was used as template for amplification of the V4 region of the 16S rRNA gene and fragments were sequenced on an Illumina MiSeq as previously described @Kozich2013. Sequences were curated using the mothur software as described previously @Schloss2009. The sequences were assigned taxonomic classification using a naive Bayesian classifier trained using a 16S rRNA gene training set from the Ribosomal Database Project (RDP) @Wang2007 and clustered into operational taxonomic units (OTUs) based on a 97% similarity cutoff. Sequencing and analysis of a mock community revealed the error rate to be 0.018%. Samples were rarefied to 4231 sequences per sample in order to reduce uneven sampling bias.

Diversity analysis was performed using the Simpson diversity calculator and $\theta$<sub>YC</sub> calculator metrics in mothur version 1.39.5 @Schloss2009. ThetaYC distances were calculated to determine the dissimilarity between two samples. Random Forest classification models were built using the randomForest R package and resultant models were used to identify the OTUs that were most important for classifying each location @Liaw2002. To get species-level information about sequences of interest, sequences were aligned using blastn and the species name was only used if the identity score was $\ge$ 99% over the full-length of the contig and matched a single reference.

>>>>

#### Random Forest classification models identify important Operational Taxonomic Units (OTUs) on each side
To identify OTUs that were distinct at each site, we constructed several Random Forest models trained using OTU relative abundances. We used 10-fold cross validation to build the first model to classify the luminal versus mucosal samples for the proximal and distal sides, independently (Figure S1A). The models performed well when classifying these samples (proximal AUC = 0.764, distal AUC = 0.908). The OTUs that were most predictive of each site were identified by their greatest mean decrease in accuracy when removed from the model. For distinguishing the proximal lumen and mucosa, OTUs affiliated with the _Bacteriodes_, _Actinomyces_, _Psuedomonas_ and _Enterobacteraceae_ were included in the best model (Figure 4A). The model to differentiate between the distal lumen and mucosa included OTUs affiliated with the _Turicibacter_, _Finegoldia_, _Peptoniphilus_ and _Anaerococcus_ (Figure 4B). These results indicated that there were fine differences between the different sites of the colon, and that these could be traced to specific OTUs on each side.

Next, we built a Random Forest model to differentiate the proximal and distal luminal samples using 10-fold cross validation. The model performed best when distinguishing the proximal versus distal mucosa (Figure S1B, AUC = 0.850) whereas the model to differentiate between the proximal versus distal lumen performed poorly (AUC = 0.580). The OTUs included in the model differentiating the distal and proximal mucosa included members of the _Porphyromonas_, _Murdochiella_, _Finegoldia_, _Anaerococcus_ and _Peptoniphilus_ (Figure 5A). The model that attempted to separate the the proximal and distal lumen included OTUs affiliated with the _Bacteroides_, _Clostridium IV_ and _Oscillibacter_ (Figure 5B). Interestingly, _Anaerococcus_ and _Finegoldia_ were distinct between the mucosa and lumen and also helped to differentiate between the proximal and distal sides.

#### Bacterial OTUs associated with CRC and IBD are found in healthy individuals 
Given that specific bacterial species have been associated with colorectal cancer and IBD, we probed our sample set for these OTUs. Among our 100 samples, the most frequent sequence associated with the _Fusobacterium_ genus was OTU179, which aligned via blastn to _Fusobacterium nucleatum subsp animalis_ (100% over full length). This is the only species of _Fusobacterium_ known to have oncogenic properties and be found on the surfaces of colorectal cancer tumors @Castellarin2011. There were 14 samples from 8 subjects with the _F. nucleatum subsp. animalis_ sequences. Of the samples with the highest relative abundance of _F. nucleatum subsp. animalis_, four of the samples were from the proximal mucosa and three from the distal mucosa (Figure S2A). The second most frequent _Fusobacterium_ sequence was OTU472, which aligned with 99% identity to _F. varium_. In addition to _F. nucleatum_, _F. varium_ has been associated with IBD @Lee2016. Four subjects harbored _F. varium_ and the samples were split evenly between the proximal and distal mucosa (Figure S2B). OTU152 was similar to the members of the _Porphyromonas_ genus and the most frequent sequence in that OTU aligned to _Porphyromonas asacharolytica_ (99% over full length), another bacterium commonly detected and isolated from colorectal tumors. OTU152 was only detected on the distal mucosa, and in fact was one of the OTUs the classification model identified as separating distal and proximal sides (Figure S2C). Among the 11 distal mucosa samples that were positive for _P. asacharolytica_, the relative abundances for this OTU ranged from 0.01% to 16%. Thus, disease-associated OTUs could be found in our sample set of 20 healthy individuals.

### Discussion
Here we identified bacterial taxa that were specific to the lumen and mucosa of the proximal and distal sides of the human colon using samples collected during an unprepared colonoscopy of healthy subjects. We found that all locations contained a range of phyla relative abundances and a range of diversity, but that there was a wide variability between subjects. Pairwise comparisons of each of the sites revealed that the proximal mucosa and lumen were most similar to each other. Further, comparison of colonoscopy-collected samples with fecal samples demonstrated that the distal lumen was most similar to feces. Random Forest models built using OTU relative abundances from each sample identified microbiota that were particular to each location of the colon. Finally, we were able to detect some bacterial OTUs associated with colonic disease in our healthy cohort. Using unprepped colonoscopies and machine learning, we have identified bacterial taxa specific to the healthy proximal and distal human colon.

When examining the relative abundance of the dominant phyla at each site (i.e. _Bacteriodes_ and _Firmicutes_), there was a wide amount of variation. This likely reflects not only the variability between human subjects, casued by differences in age, sex, diet, but also spatial "patchiness" in the gut microbiome. One study noted that the bacteria recoverable from the same mucosal sample location can be vastly different when the samples are taken just 1 cm away from each other @Hong2011. Similar patchiness was also observed in luminal contents and fecal samples themselves; there was separation of different interacting microbes along the length of a stool sample, for instance @Stearns2011. That said, across our samples, the mucosal samples harbor more _Proteobacteria_, consistent with previous studies comparing mucosal swabs to luminal content in humans @Albenberg2014. Hence, the conclusions we were able to draw from phyla analysis may have been impacted by inter-subject patchiness.

To get around the noisiness from a diverse set of samples, we built Random Forest classification models to identify the microbiota that were specific to each side and in the lumen and mucosa. For each comparison we identified the top five OTUs that were strongly predictive of one site or another. Generally, OTUs identified in each location were consistent with known physiological gradients along the gut axis @Donaldson2015. For instance, the proximal mucosa contains the highest oxygen concentrations of the colon and  harbored mucosa-associated facultative anaerobes such as _Actinomyces_ and _Enterobacteraceae_ and aerobic _Psuedomonas_. The distal mucosa was far more likely to host strictly anaerobic species such as _Porphyromonas_, _Anaerococcus_, _Finegoldia_ and _Peptoniphilus_. Thus the gut microenvironment of each location likely enriches for these specific microbiota.

In addition to identifying features that are specific to each side of the gut, the ability of the Random Forest to classify samples can serve as a proxy for similarity. That is, a higher AUC value indicates the samples are more efficiently classfied (and thus more different) than a model with a lower AUC value. For instance, the model separating the proximal and distal mucosa had an AUC of 0.850 whereas the model for classifying the proximal and distal lumen had a much lower AUC of 0.580. Further, the latter model required 44 OTUs to best separate the samples whereas the models separating the mucosa only needed 10 OTUs. The much lower AUC and need for a high number of features compared to other models suggest these locations are the most similar of the comparisons tested. We speculate that the model was less effective at classifying the proximal and distal luminal contents because the mucosal microenvironments have variable selective pressure along the colon than the luminal microenvironments.

We detected _F. nucleatum_ and _P. asacharolytica_ in 8 and 5 of our subjects, respectively. These bacteria have been shown to be predictive of colorectal cancer in humans @Baxter2016 and have oncogenic properties in cell culture and in mice @Sears2014. Interestingly, while _F. nucleatum_ was found on both sides of the colon, _P. asacharolytica_ was only detected in the distal mucosa. Not much is known about the distribution of _P. asacharolytica_ along the colon, but given its anaerobic lifestyle and asacharolytic metabolism, it is not surprising that it resides in the less-oygen-rich and protein-rich distal mucosa @Albenberg2014. In studies examining bacteria on colorectal cancer tumors, _F. nucleatum_ is more commonly detected on proximal-sided tumors, and distribution of _F. nucleatum_ decreases along the colon to rectum @Mima2016.  Of the 8 (40%) individuals positive for _F. nucleatum_ in this study, the bacterium was spread across the proximal mucosa, distal lumen and distal mucosa. Data examining bacterial biofilms on the mucosa of CRC tumors suggests that _Fusobacteria_ species are more commonly found on proximal tumors and in biofilms, indicating that it is not only the presence of the bacterium but the structure of the tumor community that contributes to _Fusobacterium's_ role in tumorigenesis @Dejea2014. Finally, _Fusobacterium_ and _Porphyromonas_ populations not only co-occur on CRC tumors but also to synergize to promote tumorigenesis in an oral cancer model @Whitmore2014 @Flynn2016. Further analysis of the distribution and activities of these pathogens along the colon is needed to elucidate a mechanism for development of CRC or IBD subtypes in the proximal or distal colon.

The _Fusobacterium_ species _nucleatum_ and _varium_ have been commonly isolated from mucosal biopsies of patients with IBD @Lee2016. Laboratory experiments with these isolates have shown that disease-isolated _F. nucleatum_ are more invasive and stimulate more TNF-$\alpha$ production than strains from healthy individuals @Strauss2011, suggesting the bacteria may increase inflammation in the gut as well @Dharmani2011. _F. varium_ isolated from UC patients caused colonic ulcers in an experimental mouse model @Ohkusa2003. _F. varium_ was only detected in three subjects and two of those samples were isolated form the proximal mucosa (Figure S2B).  _F. varium_ is most commonly isolated from UC patient biopsies from the ileum or cecum @Ohkusa2002, suggesting this species may exhibit preference for the different environmental conditions of these gastrointestinal sites. Further work will assess how gut environment may select for species which may then cause localized disease.

Specific comparisons of our findings to previously published studies of spatial variation are confounded by the use of bowel preparation methods. A rare report of a matched-colonoscopy study sampled 18 patient's colonic mucosa and luminal contents prior to and after bowel cleansing @Shobar2016. This study found that mucosal and luminal samples were distinguishable prior to bowel cleansing, but that bowel preparation resulted in an increase in shared OTUs between each site @Shobar2016. After seven days, bowel cleansing not only made the samples more difficult to distinguish, but it also decreased the diversity observed across sites. Bowel preparation clearly biases the representation of microbiota recovered from sampling the lumen or mucosa.

By revealing specific differences in microbial populations at each location in the gut via sampling an unprepared bowel, we can begin to form hypotheses about how specific host-microbe interactions can affect disease progression of proximal and distal CRC and IBD subtypes. A better understanding of microbial activities in the gut can enhance microbiome-based screening and treatment modalities for these colon diseases.

@@

#### Sample collection
At a baseline visit, subjects gave consent and were given a home collection stool kit (Zymo). At least one week prior to the scheduled colonoscopy, subjects collected whole stool at home and shipped the samples to a research coordinator on ice. Notably, subjects did not undergo any bowel preparation method prior to sampling. On the procedure day, subjects reported to the Michigan Clinical Research Unit at the University of Michigan Health System. Subjects were consciously sedated using Fentanyl, Versed and/or Benadryl as appropriate. A flexible sigmoidoscope was first inserted about 25cm into the colon and jumbo biopsy forceps used to collect the luminal contents. Two luminal samples were collected and the contents immediately deposited into RNAlater (Fischer) and flash-frozen in liquid nitrogen. The forceps were withdrawn and new biopsy forceps were used to collect mucosal biopsies on sections of the colon that were pink and free of stool matter. Three mucosal biopsies were collected and flash-frozen in RNAlater. These samples comprised the distal colon samples. The sigmoidoscope was then withdrawn and a pediatric colonoscope was inserted to reach the proximal colon. Samples were then collected in the same manner as was done in the distal colon and the colonoscope withdrawn. All samples were stored at -80$^\circ$C.

#### Sample processing, sequencing and analysis
DNA extraction was performed using the PowerMicrobiome DNA/RNA Isolation Kit (MO BIO Laboratories). For tissue biopsies, Bond-Breaker TCEP solution (Fisher) and 2.8mm ceramic beads (MO BIO Laboratories) were added to the bead beating step to enhance DNA recovery from mucosal samples. The resulting DNA was used as template for amplification of the V4 region of the 16S rRNA gene and fragments were sequenced on an Illumina MiSeq as previously described @Kozich2013. Sequences were curated using the mothur software as described previously @Schloss2009. The sequences were assigned a taxonomic classification using a naive Bayesian classifier trained using a 16S rRNA gene training set from the Ribosomal Database Project (RDP) @Wang2007 and clustered into operational taxonomic units (OTUs) based on a 97% similarity cutoff. Sequencing and analysis of a mock community revealed the error rate to be 0.018%. Samples were rarefied to 4231 sequences per sample in order to reduce uneven sampling bias.

Diversity analysis was performed using the Simpson diversity calculator and $\theta$<sub>YC</sub> calculator metrics in mothur version 1.39.5 @Schloss2009. $\theta$<sub>YC</sub> distances were calculated to determine the dissimilarity between two samples. Random Forest classification models were built using the AUCRF R package using a leave-one-subject out approach @Calle_2011. For each model the data was split into a 19-subject training set and a 1-subject test set. The model was built and cross-validated using AUCRFcv on the training set. The model was then tested on the left-out patient. This process was repeated iteratively for all subjects and results plotted as Reciever Operator Characteristic curves using the pROC R package @Robin_2011. Resultant models were used to identify the OTUs that were most important for classifying each location. Species-level information for sequences of interest was obtained by aligning the sequences to the GenBank nucleotide databse using blastn. The species name was only used if the identity score was $\ge$ 99% over the full-length of the contig and matched a single reference.
